Fast, Accurate, and Scalable Genomic Structural Variation Detection on the Cloud: MapReduce based {\em Cloudbreak}

### 2013-07-03 manuscript/nm.tex
shrink word count, more edits

\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
\author[3,4]{Jeffrey Tyner}
\author[6]{Alberto L'Abbate}
\author[6]{Clelia Tiziana Storlazzi}
\author[4,5]{Lucia Carbone}
\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
\affil[3]{Program in Molecular and Cellular Biosciences}
\affil[4]{Behavioral Neuroscience Department}
\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
\affil[6]{Department of Genetics and Microbiology, University of Bari, Bari, Italy}

@@

The detection of genomic structural variations (SV) remains one of the the most difficult challenges in analyzing high-throughput sequencing data due to the uncertainty inherent in mapping short sequencing reads and the scale of the data. The growing size and number of sequencing data sets have rendered SV detection, and high throughput sequencing analysis in general, a bona fide big data problem and led to intense interest in distributing computation to clusters of commodity servers or on-demand cloud infrastructure. Meanwhile, MapReduce is becoming a standard framework for distributed computing and has convincingly demonstrated its effectiveness in offering scalable solutions to big data problems. We describe a novel conceptual framework for SV detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion and insertion detection algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performances meeting or exceeding those of a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes, while providing dramatically shorter runtimes and the ability to readily scale up in data volume on large clusters. In particular, our algorithm excels at discovering insertions and deletions of size 40bp-100bp and can accurately genotype heterozygous and homozygous deletions from diploid samples. Finally, our software distribution includes an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, enabling on-demand cluster computing. Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.

@@

Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}. Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets and integrating them all is computationally difficult. As the volume of massively parallel sequencing data approaches scales that are associated with ``big data'', SV detection is becoming a time consuming component of genomics pipelines on traditional cluster architectures, and presents a significant challenge for genomics groups or clinical operations that may not be able to scale up their computational resources readily. Here we present a distributed software solution that is scalable and readily available on the cloud.

In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute computation to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large scale data sets across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster.  MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention. Use of the Hadoop framework has been demonstrated for several sequencing-related bioinformatic tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.

Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing analysis problems like SV detection. MapReduce divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets \textcite{Alkan:2011p547}. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment. These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

Many RP methods use only unambiguously discordantly mapped read pairs in their analyses. Some approaches also consider ambiguous mappings of discordant read pairs in an attempt to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}. Increasing the number of read mappings considered, however, increases the computational burden of SV detection. As large-scale sequencing projects and clinical sequencing applications emerge, fast runtimes and efficient storage of data are becoming increasingly important, although they have not been a focus of SV detection algorithm development.

Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines. The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features along the genome and then producing SV predictions from those features in a post-processing step.

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate the development in this framework of a software package, Cloudbreak, for discovering genomic deletions up to 25,000bp long, and short insertions. Cloudbreak computes local features that build on the ideas implemented in MoDIL \autocite{Lee:2009da}, which modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm. We characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods. Cloudbreak also provides an automatic method to set up and configure MapReduce (Hadoop) clusters on a variety of cloud services, making distributed SV detection accessible to those who wish to scale their computational infrastructure as their needs change.

@@

Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs. This framework is summarized in Figure \ref{cloudbreak_workflow}. Details of the MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{Scb_algo} and the Supplementary Materials; here we proceed with a high-level description of each job.

Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials). In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair. In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair.

In the next job, Cloudbreak computes a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping windows. For the experiments reported in this paper we use a window size of 25bp. Within each window, we examine the distribution of insert sizes of read pairs that span that window, and compute features by fitting a Gaussian Mixture Model to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}). To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.

Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location. In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls, employing noise reduction techniques to increase specificity (Supplementary Materials).  We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task. An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.

@@

We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}. DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence. We refer to these sets of calls as DELLY-RP and DELLY-SR in the following results. We also attempted to evaluate MoDIL on the same data. All of these methods detect deletions; in addition BreakDancer, Pindel, and MoDIL are capable of detecting insertions. See Methods for details on how reads were aligned and each program was invoked. For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{Salignment_comparison}).

As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J. Craig Venter \autocite{Levy:2007fb}. Using these variants, we simulated a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).

@@

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristics (ROC) curves of the performance of each algorithm for detecting deletions and insertions on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Methods for a description of how we identified correct predictions. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY. For insertions, Cloudbreak clearly provides the best combinations of sensitivity and specificity. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster, so results are not included in the runtime figure. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. Therefore, to test scalability we measured the runtime of that job on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).

Choosing the correct operating point or threshold to set on the output of an SV calling algorithm can be difficult when operating on a new data set. The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying levels. First, we characterized the predictions made by each algorithm at the operating point which gives them maximum sensitivity. For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low. The results for both deletion and insertion predictions are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}. MoDIL and Cloudbreak exhibited the greatest recall for deletions. Cloudbreak has high precision and recall for deletions at this threshold, and discovers many more small deletions. For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches. Cloudbreak again identifies the most small variants. Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by read-pair mapping. We also used the ROC curves to attempt to characterize the predictions made by each algorithm when a low false discovery rate is required. Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of simulated deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve. At this more stringent threshold, Cloudbreak identifies more deletions in every size category than any other tool. Performance on insertions never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table. We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).

It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we strive however to increase sensitivity and specificity to actual variants in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us, especially given the possibility of pipelines in which RP calls are validated \emph{in silico} by local assembly methods or split-read methods.

@@

We downloaded a data set of reads taken from a DNA sample of Yoruban individual NA18507, experiment ERX009609 from the Sequence Read Archive. This sample was sequenced on the Illumina Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage. To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same sample. These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507. We merged any overlapping calls of the same type into the region spanned by their unions. It should be noted that the 1000 Genomes call set was partially produced using DELLY and BreakDancer, and therefore those calls are ones that those tools are sensitive to, biasing this test in their favor. We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.

@@

Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm on the NA18507 data set when compared against the gold standard set for both deletions and insertions. All algorithms show far less specificity for the gold standard set than they did for the true variants in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample. For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels. Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity. Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).

Given the high number of false positives produced by all tools at maximum sensitivity, we decided to characterize the predictions made at more stringent thresholds. We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. For insertions, we were again forced to use the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data. The precision and recall at these thresholds with respect to the gold standard set, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}. For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class. Pindel exhibits the highest precision with respect to the gold standard set. For insertions, Pindel again has the highest precision at maximum sensitivity, although according the ROC curve it is possible to choose a threshold for Cloudbreak with higher precision and recall. At maximum sensitivity, Cloudbreak identifies 120 more insertions from the gold standard set than Pindel, giving it by far the highest recall.

@@

Over the next few years, due to advances in sequencing technology, genomics data are expected to increase both in volume and in coverage by several orders of magnitude, expanding into the realm referred to as Big Data. In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines. In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.

Among Big Data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters. In this paper, we introduce a novel conceptual framework for structural variation detection algorithms in MapReduce, based on computing local features along the genome. This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of a distributed deletion-finding algorithm based on fitting a GMM to the distribution of mapped insert sizes spanning each location in the genome.

>>>>

\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
\author[3,4]{Jeffrey Tyner}
\author[6]{Alberto L'Abbate}
\author[6]{Clelia Tiziana Storlazzi}
\author[4,5]{Lucia Carbone}
\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
\affil[3]{Program in Molecular and Cellular Biosciences}
\affil[4]{Behavioral Neuroscience Department}
\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}

@@

The detection of genomic structural variations (SV) remains a difficult challenge in analyzing high-throughput sequencing data, and the growing size and number of sequencing data sets have rendered SV detection a bona fide big data problem. MapReduce is a proven, scalable solution for distributed computing on huge data sets. We describe a conceptual framework for SV detection algorithms in MapReduce based on computing local genomic features, and use it to develop a deletion and insertion detection algorithm, Cloudbreak. On simulated and real data sets, Cloudbreak achieves accuracy improvements over popular SV detection algorithms across a range of variant sizes, especially those 40bp-100bp. Cloudbreak can accurately genotype variants from diploid samples. On large clusters, it provides dramatically shorter runtimes and the ability to scale to big data volumes. Cloudbreak includes tools to set up and configure MapReduce (Hadoop) clusters on cloud services, enabling on-demand cluster computing. Our implementation and source code are available at \url{http://github.com/cwhelan/cloudbreak}.

@@

Genomic structural variations (SVs) such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of SVs with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}. Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them all is computationally difficult. As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure. Here we present a distributed software solution that is scalable and readily available on the cloud.

In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce \autocite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers. Hadoop is an open source project of the Apache Foundation which provides an implementation of MapReduce. MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention. The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \autocite{Schatz:2009p278}, SNP calling \autocite{Langmead:2009p1225}, and RNA-seq differential expression analysis \autocite{Langmead:2010p1268}.

Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection. MapReduce divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets \textcite{Alkan:2011p547}. Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

Many RP methods consider only unambiguously discordantly mapped read pairs. Some approaches also include ambiguous mappings of discordant read pairs to improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. Several recent RP approaches have also considered concordant read pairs, either to integrate RD signals for improved accuracy \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds that separate concordant from discordant mappings and thus detect smaller events \autocite{Marschall:2012ek}. Increasing the number of read mappings considered, however, increases the computational burden of SV detection.

Our goal is to leverage the strengths of the MapReduce computational framework in order to provide fast, accurate and readily scalable SV-detection pipelines. The main challenge in this endeavor is the need to separate logic into mappers and reducers, which makes it difficult to implement traditional RP-based SV detection approaches in MapReduce, particularly given the global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In sequencing applications, for example, MapReduce based SNV-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations on partitions of the genome. SV approaches that are similarly based on local computations have been described: the RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} compute scores or features along the genome and then produce SV predictions from those features in a post-processing step. We will show that this strategy can be translated into the MapReduce architecture.

In this paper, we describe a framework for solving SV detection problems in Hadoop based on the computation of local genomic features from paired end mappings. In this framework we have developed a software package, Cloudbreak, that discovers genomic deletions up to 25,000bp long, and short insertions. Cloudbreak computes local features based on modeling the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM), an idea first implemented in MoDIL \autocite{Lee:2009da}. The use of Hadoop enables scalable implementations of this class of algorithm. We characterize our algorithm's performance on simulated and real data sets and compare it to several popular methods. Our implementation of Cloudbreak also provides logic to set up and configure Hadoop clusters on cloud service providers, making distributed SV detection accessible to those who wish to dynamically scale their computational infrastructure as their needs change.

@@

Our framework for SV detection in MapReduce divides processing into three distinct MapReduce jobs (Figure \ref{cloudbreak_workflow}): a job that can align reads to the reference using a variety of mapping algorithms; a job that computes a set of features along the genome; and a job which calls structural variations based on those features. Our description of a general MapReduce SV detection algorithmic framework and how Cloudbreak is implemented within that framework are provided in Supplementary Algorithm \ref{Scb_algo} and the Supplementary Materials; here we proceed with a high-level description of the Cloudbreak implementation.

Cloudbreak's alignment job can run a variety of alignment tools that report reads in SAM format (Supplementary Materials). In the map phase, mappers align reads in either single-end or paired-end mode to the reference genome in parallel, outputting mapping locations as values under a key identifying the read pair. In the reduce phase, the framework combines the reported mapping locations for the two ends of each read pair. This job can also be skipped in favor of importing a pre-existing set of mappings directly into the Hadoop cluster.

In the next job, Cloudbreak computes a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping windows. For the experiments reported in this paper we use a window size of 25bp. Within each window, we examine the distribution of insert sizes of mappings that span that window, and compute features by fitting a GMM to that distribution (Supplementary Materials, Supplementary Figure \ref{Sinsert_size_mixes}). To remove incorrect mappings, we use an adaptive quality cutoff for each genomic location and then perform an outlier-detection based noise reduction technique (Supplementary Materials); these procedures also allow us to process multiple mappings for each read if they are reported by the aligner.

Finally, the third MapReduce job is responsible for making SV calls based on these features. In this job, we search for contiguous blocks of genomic locations with similar features and merge them into individual insertion and deletion SV calls after applying noise reduction (Supplementary Materials).  Reducers process each chromosome in parallel after mappers find and organize its features. An illustration of the Cloudbreak algorithm working on a simple example is shown in Supplementary Figure \ref{Salgorithm_example}.

@@

We compared the performance of Cloudbreak for detecting deletions and insertions to a selection of popular tools: the RP method BreakDancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}. DELLY produces two sets of calls, one based solely on RP signals, and the other based on RP calls that could be supported by SR evidence; we refer to these sets of calls as DELLY-RP and DELLY-SR. We also attempted to evaluate MoDIL on the same data. All of these methods detect deletions. Insertions can be detected by BreakDancer, Pindel, and MoDIL. See Methods for details on how reads were aligned and each program was invoked. For all alignments we used BWA \autocite{Li:2009p836}, although in testing Cloudbreak we have found that the choice of aligner, number of possible mapping locations reported, and whether the reads were aligned in paired-end or single-ended mode can have a variety of effects on the output of the algorithm (Supplementary Figure \ref{Salignment_comparison}).

There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of SVs. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, any simulated data should contain realistic SVs that follow patterns observed in real data. Therefore, we took one of the most complete lists of SVs from an individual, the list of homozygous insertions and deletions from the genome of J. Craig Venter \autocite{Levy:2007fb}, and used it to simulate a 30X read coverage data set for a diploid human Chromosome 2 with a mix of homozygous and heterozygous variants, with 100bp reads and a mean fragment size of 300bp (See Methods).

@@

Figure \ref{chr2CombinedRoc} shows Receiver Operating Characteristic (ROC) curves of the performance of each algorithm at identifying variants larger than 40bp on the simulated data set, as well as the runtimes of the approaches tested, excluding alignment. See Methods for a description of how we identified correct predictions. All approaches show excellent specificity at high thresholds in this simulation. Cloudbreak provides the greatest specificity for deletions at higher levels of sensitivity, followed by DELLY. For insertions, Cloudbreak's performance is similar to or slightly better than Pindel. Cloudbreak's runtime is half that of BreakDancer, the next fastest tool, processing the simulated data in under six minutes. (Of course, Cloudbreak uses many more CPUs as distributed algorithm. See Supplementary Material and Supplementary Table \ref{Sruntimes} for a discussion of runtimes and parallelization.) The output which we obtained from MoDIL did not have a threshold that could be varied to correlate with the trade-off between precision and recall and therefore it is not included in ROC curves; in addition, MoDIL ran for 52,547 seconds using 250 CPUs in our cluster. Apart from the alignment phase, which is embarrassingly parallel, the feature generation job is the most computationally intensive part of the Cloudbreak workflow. To test its scalability we measured its runtime on Hadoop clusters made up of varying numbers of nodes and observed that linear speedups can be achieved in this portion of the algorithm by adding additional nodes to the cluster until a point of diminishing returns is reached (Supplementary Figure \ref{Sscalability}).

Choosing the correct threshold to set on the output of an SV calling algorithm can be difficult. The use of simulated data and ROC curves allows for some investigation of the performance characteristics of algorithms at varying thresholds. First, we characterized the predictions made by each algorithm at the threshold that gives them maximum sensitivity. For Cloudbreak we chose an operating point at which marginal improvements in sensitivity became very low. The results are summarized in Table \ref{chr2DeletionAndInsertionPredsMaxSensitivity}. MoDIL and Cloudbreak exhibited the greatest recall for deletions. Cloudbreak has also has high precision at this threshold, and discovers many small variants. For insertions, Cloudbreak has the highest recall, although recall is low for all four approaches. Cloudbreak again identifies many small variants. Pindel is the only tool which can consistently identify large insertions, as insertions larger than the library insert size do not produce mapping signatures detectable by RP mapping. We also used the ROC curves to characterize algorithm performance when a low false discovery rate is required. Supplementary Table \ref{Schr2DeletionPredsFDR10} shows the total number of deletions found by each tool when choosing a threshold that gives an FDR closest to 10\% based on the ROC curve. At this more stringent cutoff, Cloudbreak identifies more deletions in every size category than any other tool. Insertions performance never reached an FDR of 10\% for any threshold, so insertion predictions are not included in this table. We also examined Cloudbreak's ability to detect events in repetitive regions of the genome, and found that it was similar to the other methods tested (Supplementary Tables \ref{SdeletionRepmaskpreds} and \ref{SinsertionRepmaskpreds}).

It should be noted that the methods tested here vary in their breakpoint resolution (Supplementary Figure \ref{Sbreakpoint_resolution}): SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution by dividing the genome into 25bp windows; we believe, however, that increasing sensitivity and specificity is of greatest utility, especially given the emergence of pipelines in which RP calls are validated \emph{in silico} by local assembly.

@@

We downloaded a set of reads from Yoruban individual NA18507, experiment ERX009609, from the Sequence Read Archive. This sample was sequenced by Illumina Inc. on the Genome Analyzer II platform with 100bp paired end reads and a mean fragment size (minus adapters) of 300bp, with a standard deviation of 15bp, to a depth of approximately 37X coverage. To create a gold standard set of insertions and deletions to test against, we pooled annotated variants discovered by three previous studies on the same individual. These included data from the Human Genome Structural Variation Project reported by \textcite{Kidd:2008p926}, a survey of small indels conducted by \textcite{Mills:2011fi}, and insertions and deletions from the merged call set of the phase 1 release of the 1000 Genomes Project \autocite{GenomesProjectConsortium:2012co} which were genotyped as present in NA18507. We merged any overlapping calls of the same type into the region spanned by their unions. We were unable to run MoDIL on the whole-genome data set due to the estimated runtime and storage requirements.

@@

Figure \ref{NA18507CombinedRoc} shows the performance of each algorithm at detecting events larger than 40bp on the NA18507 data set. All algorithms show far less specificity for the gold standard set than they did in the single chromosome simulation, although it is difficult to tell how much of the difference is due to the added complexity of real data and a whole genome, and how much is due to missing variants in the gold standard set that are actually present in the sample. For deletions, Cloudbreak is the best performer at the most stringent thresholds, and has the highest or second highest precision at higher sensitivity levels. Cloudbreak has comparable accuracy for insertions to other tools, and can identify the most variants at higher levels of sensitivity. Cloudbreak processes the sample in under 15 minutes on our cluster, more than six times as fast as the next fastest program, BreakDancer, even when BreakDancer is run in parallel for each chromosome on different nodes in the cluster (see Methods and Supplementary Material).

Given the high number of novel predictions made by all tools at maximum sensitivity, we decided to characterize performance at more stringent thresholds. We examined the deletion predictions made by each algorithm using the same cutoffs that yielded a 10\% FDR on the simulated chromosome 2 data set, adjusted proportionally for the difference in coverage from 30X to 37X. For insertions, we used the maximum sensitivity thresholds for each tool due to the high observed FDRs in the simulated data. Precision and recall at these thresholds, as well as the performance of each algorithm at predicting variants of each size class, is shown in Table \ref{NA18507DeletionAndInsertionPreds}. For deletions, Cloudbreak has the greatest sensitivity of any tool, identifying the most variants in each size class. Pindel exhibits the highest precision with respect to the gold standard set. For insertions, Pindel again has the highest precision at maximum sensitivity, while Cloudbreak has by far the highest recall.

@@

Over the next few years, due to advances in sequencing technology, genomics data are expected to increase in volume by several orders of magnitude, expanding into the realm referred to as ``big data''. In addition, the usage of existing information will increase drastically as research in genomics grows and translational applications are developed, and the data sets are reprocessed and integrated into new pipelines. In order to capitalize on this emerging wealth of genome data, novel computational solutions that are capable of scaling with the increasing number and size of these data sets will have to be developed.

Among big data infrastructures, MapReduce is emerging as a standard framework for distributing computation across compute clusters. In this paper, we introduced a novel conceptual framework for SV detection algorithms in MapReduce, based on computing local genomic features. This framework provides a scalable basis for developing SV detection algorithms, as demonstrated by our development of an algorithm for detecting insertions and deletions based on fitting a GMM to the distribution of mapped insert sizes spanning each genomic location.
